A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-09 analysis/paper/paper.Rmd
update paper

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`). We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "4-method_characterisation")`).

@@

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "4-method_characterisation", "a")`). However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

`r anchor("fig", "4-method_characterisation")`
![](`r figure_file("method_characteristics.svg", experiment_id="4-method_characterisation")`)
`r ref("fig", "4-method_characterisation", anchor=TRUE)` First characterization of each method.

@@

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "4-method_characterisation", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "4-method_characterisation", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "4-method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

We wrapped each method and postprocessed its output into a common probabilistic format. This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2

# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focussing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

>>>>

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface (`r ref("table", "methods")`). We characterized all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characterisation")`).

@@

At the level of trajectory topologies, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characterisation", "a")`). However, only a handful, mostly recent, methods can currently handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

`r anchor("fig", "method_characterisation")`
![](`r figure_file("method_characteristics.svg", experiment_id="method_characterisation")`)
`r ref("fig", "method_characterisation", anchor=TRUE)` First characterization of each method.

@@

Most methods fulfill most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characterisation", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characterisation", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characterisation", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

We wrapped each method and postprocessed its output into a common probabilistic format. This format works at two level: at the To evaluate the accuracy of each method, we gathered both synthetic and real datasets, and used these to assess how good a method is at correctly ordering cells, detecting the right trajectory topology and 2
# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] created by Sean Davis and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focussing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.
